---
title: "Causal analysis of the relationship between mastitis and longevity in dairy cows"
author: "Aurélien Madouasse"
format: html
bibliography: references.bib
---

```{r setup}
library(tidyverse)
library(ggdag)
library(DairyHerdData)
library(ggplot2)
library(scales)
library(questionr)
library(gmodels)
library(lme4)
```

# Introduction

In broad terms, longevity is the length of the life of an animal, although a wide variety of indicators can be used to measure it. A recent study by @dallago2021 provided evidence of a decrease in dairy cows' longevity between the 1960s and 2010s in the top 10 high milk-producing countries.

The departure of a cow from her herd is called culling [@fetrow2006]. In some cases, the farmer has no other choice than to cull a cow, in case of death or when the cow is not pregnant and does not give enough milk anymore to be kept in the herd. This is usually called involuntary culling. In other cases, farmers choose to cull the cows for various reasons, mostly for the presence of a disease. This is called voluntary culling. The main documented causes of culling are reproductive failure, mastitis and lameness.

Improving dairy cows' longevity therefore requires to act on the causes of culling. For example, if a disease is reported as a cause for culling, preventing the disease from occurring should increase longevity. However, this is a notably hard task. The decision of culling a cow may depend on several combined causes that are related to the status of a given animal as well as on the farm context. Furthermore, disease incidence may be increased by the presence of risk factors, and, disease occurrence can also act on factors such as milk production or reproduction that are sufficient reasons for culling. Lastly, for contagious diseases, culling infected cows could reduce the incidence of the disease by removing sources of infection from the herd.

Our objective in this paper is to analyse longevity and disease occurrence in a causal framework, using mastitis in dairy cows as an example.

# Introduction

## Causal framework

Determining what causes what is an important question in many fields. Philosophers and scientists have come up with various ways of addressing it. In quantitative disciplines such as epidemiology, economics and sociology, the 2 paradigms currently in use are the *potential outcomes framework* who is associated with Donald Rubin [@rubin1974] and *structural causal models* proposed by Judea Pearl [@pearl2009]. Although the 2 paradigms seem to differ in important ways [@markus2021], they are sometimes presented as part of a single causality framework in textbooks [@neal2020; @cunningham2021]. In what follows, we frame our problem using the potential outcomes framework and then decompose it using directed acyclic graphs, most commonly associated with the structural causal framework. Our objectives are to make sound causal inference while keeping the presentation of our research questions to the readers as clear as possible.

## Potential outcomes framework

Estimating the causal effect of the occurrence of an event on an outcome requires to be able to compare the outcome values with and without the event. For example, estimating the effect of mastitis occurrence on the probability of culling in a given cow would require to be able to observe the same cow with and without mastitis and record whether she is culled or not following the disease or its absence. Such data is of course impossible to obtain. This is in fact called *the fundamental problem of causal inference*. But laying out the problem in these terms can help to clarify the research questions, design experiments, analyse the data and interpret the results. @hernan2024 describe in simple terms and introduce notation for the potential outcome framework, that was developed by several scientists over the course of the XX$^{th}$ century, notably Donald Rubin [@rubin1974]. Let us introduce this notation using the causal effect of mastitis on the probability of culling in a cow as an example. We consider mastitis occurrence as a binary event $A$ with $A=1$ denoting presence of mastitis and $A=0$ denoting the absence of mastitis. We are interested in what would happen in cow $i$ if she were to experience mastitis ($Y_i^{1}$) or, on the contrary, if she weren't experiencing mastitis ($Y_i^{0}$). $Y_i^{A}$ represents whether cow $i$ is culled or not at the end of her lactation, with $Y^{A=a}_i=1$ if cow $i$ is culled and $Y^{A=a}_i=0$ if cow $i$ is not culled, when event $A=a$ occurs. In a given cow, only one event $A=a$ and its associated outcome $Y^{A=a}$ can be observed. These event and outcome are called factual, and the unobserved ones are called counterfactual. The factual (realised) outcomes are encoded using the $|$ symbol. For example, $Y_i | A = 0$ is the observed outcome for cow $i$ who did not get mastitis. Problems arise when there are systematic differences in potential outcomes between treated and untreated individuals, for example if $Y^{A = 1}|A = 1 \neq Y^{A = 1}|A = 0$.

## Estimands: the quantities of interest

Estimands are quantities whose value is the answer to a precise research question. Estimands are independent of both the data available and the estimation method used for the analysis. They are used in slightly different ways in different disciplines (see for example the different perspectives in sociology in @lundberg2021a and in randomized clinical trials in @kahan2024 ). Following @lundberg2021a, the two minimum components of an estimand are a *unit-specific quantity* and a *target population* [@lundberg2021]. These authors make a further distinction between theoretical estimands which are the quantities of interest to the researchers and empirical estimands which are approximations that can be computed from observed data.

For example, if we wanted to know the causal effect of clinical mastitis in early lactation on the probability of culling before or at the end of lactation in French dairy cows, we could define the following theoretical estimand:

$$
\delta_1 = \frac{1}{n} \sum_{i=1}^n Y^{1}_i - Y^{0}_i
$$ where $n$ is the total number of cows in the French dairy cow population, $Y^1_i$ and $Y^0_i$ are binary variables taking the value 1 for cows culled at or before the end of lactation and 0 otherwise in cows with ($Y^1_i$ ) and without ($Y^0_i$) clinical mastitis in early lactation respectively. Of course, in each cow, only one of the 2 outcomes can be observed. This estimand is a risk difference, but risk ratios or odds ratios could have been used as well.

But maybe, the cows who got mastitis are different from the cows who did not, for reasons other than having experienced the disease, and therefore have a different probability of being culled. $Y^{A=a}_i|A=1$ denotes the outcome associated with having mastitis given that animal $i$ experienced mastitis. We could make the hypothesis that $Y^1_i|A=1 \neq Y^1_i|A=0$ and choose to focus our interest on cows who actually got mastitis. A theoretical estimand of interest in this situation could be: $$
\delta_2 = \frac{1}{n} \sum_{i=1}^n \left[ (Y^{1}_i|A=1) - (Y^{0}_i|A=1) \right]
$$

$\delta_1$ is called the average treatment effect (**ATE**). It is a measure of a treatment (in our case a disease) effect in the whole population. $\delta_2$ is called the average treatment effect in the treated (**ATT**). In our case, it is more relevant to consider the ATE, since our interest is in evaluating the consequences of preventing mastitis in cows who experienced the disease, not really on the whole population.

## Directed acyclic graphs

Directed acyclic graphs, commonly called **DAG**s, are graphical representations of a phenomenon which are particularly well suited to model causality. The different parts or components of the phenomenon are represented using circles called nodes. We can represent the hypothesis that mastitis ($M$) causes culling ($C$) as follows:

```{r dag1}
#| echo: false
#| fig-width: 3
#| fig-height: 1
dag1 <- dagify(
  C ~ M,
  coords = list(
    x = c(M = 0, C = 2),
    y = c(M = 0, C = 0)
  )
)

ggdag(dag1) + theme_dag()
```

where the arrow indicates causality. DAGs allow to represent complex relationships between variables. We now provide a few examples of how DAGs can be useful to get a better picture of causal relationship.

Mastitis could have both direct and indirect effects on culling. As a direct effect, a severe case of clinical mastitis can lead to the death of the cow. As an indirect effect, mastitis could compromise reproductive success. This indirect effect of mastitis is *mediated* by reproduction. One usual way of modelling such data would be to fit a logistic model with culling as the outcome and mastitis and reproduction as covariates. However, if there is an effect of mastitis on reproduction, the true causal effect of mastitis on culling by the logistic model will be underestimated, because part of if will be incorporated into the effect of reproduction on culling.

```{r dag2}
#| echo: false
#| fig-width: 3
#| fig-height: 3
dag2 <- dagify(
  C ~ M,
  R ~ M,
  C ~ R,
  coords = list(
    x = c(M = 0, C = 2, R = 1),
    y = c(M = 0, C = 0, R = -1)
  )
)

ggdag(dag2) + theme_dag()
```

## Definition of mastitis

Mastitis is an inflammation of a mammary gland, caused by an intramammary infection. Different aetiological agents are associated with differences in disease epidemiology. On the one hand, bacteria of faecal origin such as *Escherichia coli* cause short lived infections with a high rate of spontaneous cure, but often acute clinical signs. On the other hand, some bacteria, such as *Staphylococcus aureus* are adapted to survive in the mammary gland and are associated with chronic infections. In this case, the infection can last for months, with clinical signs observed on rare occasions with regards with the duration of the infection. Because bacteria are shed in the milk of infected cows for long periods, milk is more likely to serve as a reservoir of infection. These infections are therefore described as contagious. All infections result in the recruitment of immune cells in the mammary gland, that are measured on a monthly basis in herds that participate to milk recording. The measured cell concentration is called milk somatic count (**SCC**).

In this study, we chose to consider the following types of mastitis:

-   new case of contagious mastitis in early lactation: primiparous cow with at least 2 out of the first 3 milk recordings above 200 000 cells/mL or multiparous cow with the last milk recording in her previous lactation below 200 000 cells/mL and at least 2 out of the first 3 milk recordings above 200 000 cells/mL
-   environmental mastitis in early lactation

## Definition of culling

## Hypohteses and DAG

# Data analysis

Unité statistique = lactation Y = réforme \<-\> revêlage

Using the `DairyHerdData` package.

```{r milk-recording-ai-data}
## milk recording data
rec <- rec
## artificial insemination data
ai <- ai |> 
  mutate(ai_date = as.Date(ai_date))

## days in milk added to milk recording data
rec <- rec |> 
  mutate(dim = as.integer(ctrl_date - calv_date))

## making a list of lactations
lac <- rec |> 
  group_by(anim_id, calv_date) |> 
  summarise(lac_id = NA,
            parity = unique(parity),
            n_ctrl = length(unique(ctrl_date)),
            n_scc_high = length(scc[scc > 200]),
            n_scc_early = length(scc[dim < 91]),
            n_scc_early_high = length(scc[scc > 200 & dim < 91]),
            n_scc_late_high = length(scc[scc > 200 & dim > 91]),
            #rajout de la moyenne de la PL les 91 premiers jours de lactation
            early_plmoy = mean(milk[dim < 91])) |> 
  ungroup(anim_id)

lac$lac_id <- 1:nrow(lac)

lac <- lac |> 
  group_by(anim_id) |> 
  mutate(calv_nxt = lead(calv_date))

## first date of calving
calv_first <- min(lac$calv_date)

ai <- ai |> 
  filter(ai_date > calv_first)

ai1 <- left_join(lac |> 
                   select(anim_id, lac_id, calv_date, calv_nxt), 
                 ai |> 
                   select(herd_id, cow_id, ai_date),
                 join_by(anim_id == cow_id)) |> 
           filter(ai_date > calv_date & (ai_date < calv_nxt | is.na(calv_nxt))) |> 
  select(herd_id, anim_id, lac_id, calv_date, calv_nxt, ai_date)

ai_lac <- ai1 |> 
  arrange(lac_id, ai_date) |> 
  group_by(lac_id) |> 
  summarise(ai_N = length(unique(ai_date)),
            ai_first = min(ai_date),
            ai_last = max(ai_date),
            ai_calv_days = as.integer(unique(calv_nxt) - max(ai_date)))

lac <- left_join(lac, ai_lac)

```

# References

# Essai Joanna

@Joanna : 
- Je mets un eval = false sur certains chunks. De cette manière, les chunks ne sont pas évalués quand tu exécutes le fichier ('render').
- Il est recommandé de nommer les chunks pour faciliter la lecture quand on extrait le code R.
- Eviter de modifier les jeux de données par les numéros de colonnes. C'est peu lisible et ça favorise les erreurs.
- Charger les packages en début de code. Encore une fois, c'est plus lisbile. Ca évite de les charger plusieurs fois.



Etablissons le lien entre mammites et mise à la reproduction (décision prise par l'éleveur). n_scc_early_high donne le nombre de contrôles pour lesquelles le taux de cellules est supérieur à 200 000 dans les 91 premiers jours de lactation. On va donc émettre l'hypothèse que si n \>= 2 alors la vache à fait un épisode de mammite chronique et ce qu'on cherche à savoir c'est si ce résultat à un impact sur la décision par l'éleveur d'inséminer sa vache.

## création d'une colonne miserepro qui donne la valeur 1 si la vache à été inséminée et qu'elle présente un n_scc_early_high \>= 2


-   Une variable `miserepro` est créée. Elle vaut 1 si le nombre d'IA est supérieur à 0 ce qui signe une décision de l'élveur de mettre la vache à la reproduction.
- une variable `early_mamchro`  est créée. Elle vaut 1 pour les lactations qui ont plus de 2 CCS supérieures à 200 000 cellules sur les 3 premiers contrôles du début de lactation.

```{r miserepro-early_mamchro-definition}
lac <- lac |> 
  mutate(miserepro = ifelse(ai_N == 0 | is.na(ai_N), 0, 1),
         early_mamchro = ifelse(n_scc_early_high >= 2, 1, 0))
```

## création d'un nouveau tableau avec que les animaux atteints de mammites chronique dans les premiers 91 jours de lactation

Je pense que tu veux le tableau suivant :
--> il y a un problème avec ce tableau les valeurs de mise à la repro ne correspondent pas à celles du tableau lac

```{r}
#comparaison de la mise à la reproduction selon la présence ou non d'une mammite chronique en début de lactation
tab_repro_mam <- lac |> 
  select(miserepro, early_mamchro) |> 
  table()
  head(tab_repro_mam)
```
je me suis permise d'en créer un nouveau pour refaire un essai : 


```{r}
tab_repro_mam_1 <- lac |> 
  select(miserepro,early_mamchro) |> 
  mutate(miserepro = as.factor(miserepro), early_mamchro = as.factor(early_mamchro))
```

```{r}
p<-ggplot(tab_repro_mam_1) +
aes(x = early_mamchro, fill = miserepro) +
geom_bar(position="fill") +
xlab("mise à la repro selon mammite chronique ou non") +
ylab("Proportion") +
labs(fill = "miserepro")
p+theme_bw()
```


Ci-dessous, les proportions de vaches mises à la repro en fonction de la présence d'une mammite chronique. La relation est conforme à ce qui est attendu. Utilisation d'une fonction apply. Je pourrai t'expliquer comment ça fonctionne.

--> je ne comprend pas ce que ce chunk est censé faire. Si j'ai bien compris la fonction apply() applique la même fonction a tout les élements d'une ligne ou d'une colonne. Mais pourquoi faire la proportion de vaches mises à la repro pour chaque colonne ? 

```{r}
apply(tab_repro_mam, 2, function(x) round(x / sum(x), 5))
```


On remarque que sur les vaches qui présentent une mammite chronique précoce, 74% des vaches sont quand même mises à la reproduction.

on veut savoir s'il y a un lien entre la présence ou non de mammite et la mise à la reproduction (attention on ne met pas en évidence une relation de cause à effet)


```{r}
#
tab1 <- table(lac$miserepro, lac$early_mamchro)
tab1
lprop(tab1)
```

```{r}
CrossTable(lac$miserepro, lac$early_mamchro, digits=3, max.width = 5,
expected=FALSE, prop.r=TRUE, prop.c=TRUE,
prop.t=TRUE, prop.chisq=FALSE, chisq = FALSE, fisher=FALSE, mcnemar=FALSE,
resid=FALSE, sresid=FALSE, asresid=FALSE,
missing.include=FALSE, format=c("SAS","SPSS"))
```

```{r}
CrossTable(lac$early_mamchro,lac$miserepro, digits=3, max.width = 5,
prop.t=FALSE, prop.chisq=FALSE, chisq = TRUE, fisher=FALSE, mcnemar=FALSE,
expected=TRUE, prop.r=FALSE, prop.c=FALSE,
resid=FALSE, sresid=FALSE, asresid=FALSE,
missing.include=FALSE, format=c("SAS","SPSS"))
```

p-value \< 0,05, il existe donc un lien entre la présence d'une mammite chronique et la mise à la reproduction....

## régression logistique pour identifier la présence ou non de mammite comme facteur de risque à la non mise à la reproduction

```{r}
# première tentative de régression logistique
myreg <- glm(as.numeric(miserepro)~ early_mamchro, family= binomial, data = lac)
summary(myreg)
```

```{r simple-log-reg}
myreg <- glm(miserepro ~ early_mamchro, family= binomial(link = "logit"), data = lac)
summary(myreg)
```

On peut regarder l'odds-ratio de la mise à la repro en fonction du statut.

```{r log-reg-OR-1}
## coefficients
coef(myreg)

## OR
exp(coef(myreg)[1])
```
Version mixed de ce modèle avec effet aléatoire animal. Pas forcément pertinent. Idéalement, il faudrait mettre l'élevage en effet aléatoire. La colonne correspondante devra être ajoutée.

```{r mixed-log-reg}
myreg_mixed <- glmer(miserepro ~ early_mamchro + (1 | anim_id) , 
                    family= binomial(link = "logit"), data = lac)
summary(myreg_mixed)
```





Ce résultat nous indique que le choix de la mise à la reproduction est influencé par la présence ou non de mammite lors des 91 premiers jours de lactation. Or pour ce premier modèle j'ai totalement ignoré les autres variables qui peuvent influencer ce résultat comme : - la parité - la production laitière - la présence d'une autre affection concomitante


rajout d'une nouvelle variable à ce modèle : prenons le rang de lactation 

```{r}
#rajout de la variable parité dans la régression logistique 
myreg2 <- glm(miserepro ~ early_mamchro + parity, family= binomial(link = "logit"), data = lac)
summary(myreg2)
```

Le coefficient semble logique, plus la parité augmente et moins la probabilité de mise à la repro est élevée. On voit également que la valeur est très significative 

Il faudrait égalemement ajouter la production laitière (simplement la production en kg sans prendre en compte le TB et le TP)
le problème c'est que la production laitière se trouve dans le data frame rec qui ne se présente pas de la même manière. 
Il faudrait peut-être calculer la moyenne de la PL sur l'ensemble des contrôles fait avant 91j de lactation et se servir de la moyenne dans le calcul ? 

```{r}
#Création d'un nouveau jeu de données qui donne la production laitière moyenne par lactation
rec_plmoy_parite <- rec |> 
  group_by(anim_id, parity) |> 
  summarize(plmoy = mean(milk, na.rm = TRUE),
            herd_id = unique(herd_id))
```

```{r}
lacpl <- inner_join(lac |> 
                     select(anim_id, parity, miserepro, early_mamchro, early_plmoy), 
                   rec_plmoy_parite |> 
                     select(anim_id, herd_id, parity, plmoy),
                   join_by(anim_id, parity)) |> 
  select(herd_id, anim_id, parity, early_plmoy, plmoy, early_mamchro, miserepro)
  
```

```{r}
myreg3 <- glm(miserepro ~ early_mamchro +factor(parity) + plmoy, family= binomial(link = "logit"), data = lacpl)
summary(myreg3)
```

```{r}
#nouvelle régression mais avec early_plmoy 
myreg4 <- glm(miserepro ~ early_mamchro +factor(parity) + early_plmoy, family= binomial(link = "logit"), data = lacpl)
summary(myreg4)
```


ajout de la variable aléatoire de troupeau

```{r}
myreg_mixed2 <- glmer(miserepro ~ early_mamchro + factor(parity) + early_plmoy + (1 | herd_id), 
                    family= binomial(link = "logit"), data = lacpl)
summary(myreg_mixed2)
```

Regroupement des parités selon des classes plus pertinentes : On voit que sur la régression linéaire, le résultat est significatif jusqu'a une parité de 10 on peut donc regrouper les classes de parité 9 et plus. 

```{r}
#Création d'une nouvelle variable parité en 9 classes
lacpl <- lacpl |> 
  mutate(C9parite = case_when(parity == 1 ~ "1",
                            parity == 2 ~ "2",
                            parity == 3 ~ "3",
                            parity == 4 ~ "4",
                            parity == 5 ~ "5",
                            parity == 6 ~ "6",
                            parity == 7 ~ "7",
                            parity == 8 ~ "8",
                            parity >= 9 ~ "9 et plus",
                            TRUE~NA))
```
```{r}
#Régression logistique finale 2 pour la décision de mise à la repro
myreg_mixed3 <- glmer(miserepro ~ early_mamchro + factor(C9parite) + early_plmoy + (1 | herd_id), 
                    family= binomial(link = "logit"), data = lacpl)
summary(myreg_mixed3)
```

# Evaluation de l'impact des mammites sur la mort de l'animal 

Dans le tableau "mvt" on considère M = mort et E = équarissage qui est équivalent à mort

on considère cette fois 3 types de mammites : 
-mammite chronique de début de lactation (2 ctrl scc > 200 000)
-mammite chronique de fin de lactation (idem)
-mammite aiguë (1 scc < 200 000 suivi d'un scc > 200 000 et a nouveau suivi d'un scc < 200 000)

```{r}
str(rec)
```


```{r}
rec <- rec |> 
  group_by(anim_id, parity) 
```

```{r}
rec$scc_aig = rep(NA, nrow(rec))
```


```{r}
rec <- rec[order(rec$anim_id, rec$parity),]
```


```{r}
for(i in 2 : (nrow(rec)-1)){
    rec$scc_aig[i] = ifelse(rec$scc[i] >= 200
           & rec$scc[i-1] < 200
           & rec$scc[i+1] < 200
           & rec$anim_id[i-1] == rec$anim_id[i]
           & rec$anim_id[i] == rec$anim_id[i+1]
           & rec$parity[i-1] == rec$parity[i]
           & rec$parity[i] == rec$parity[i+1],
           1,
           0)
    }
```


```{r}

  
```

